---
title: "Basic Trends in Income Support Trends"
author: 
  - "Andriy Koval"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
    code_folding: show
    theme: simplex
    highlight: tango
editor_options: 
  chunk_output_type: console
---

This report explores data on Income Support program taken from the [open.alberta.ca](https://open.alberta.ca/opendata/income-support-caseload-alberta) portal.

<!--  Set the working directory to the repository's base directory; this assumes the report is nested inside of two directories.-->
```{r, echo=F, message=F, results = 'hide'}
```


```{r set_options, echo=F}
# cat("Working directory: ", getwd()) # turn on to test the location
report_render_start_time <- Sys.time()
# set options shared by all chunks
knitr::opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  dpi        = 400, # dots per inch,
  out.width  = "900px", # pixels, this affects only the markdown, not the underlying png file.  The height will be scaled appropriately.
  fig.width    = 11, # inches
  fig.height   = 6, # inches
  out.width    = 1200,
  fig.path     = 'figure-png-com/' # where figures are stored
)
echo_chunks    <- FALSE #Toggle for debugging.
message_chunks <- FALSE #Toggle for debugging.
options(width=100) # number of characters to display in the output (dflt = 80)
ggplot2::theme_set(ggplot2::theme_bw()) # common theme for all graphs
```


<!-- Load the sources.  Suppress the output when loading sources. --> 
```{r load-sources, echo = F}
source("../../scripts/common-functions.R")
```


<!-- Load packages, or at least verify they're available on the local machine.  Suppress the output when loading packages. -->
```{r load-packages, message=message_chunks, echo =F}
library(magrittr)  # pipes
library(dplyr)     # data wrangling
library(ggplot2)   # graphs
library(janitor)   # tidy data
library(tidyr)     # data wrangling
library(forcats)   # factors
library(stringr)   # strings
library(lubridate) # dates
```


<!-- Load any Global functions and variables declared in the R file.  Suppress the output. --> 
```{r declare-globals, echo =F}
# custom function for HTML tables
neat <- function(x, output_format = "html"){ 
  # knitr.table.format = output_format
  if(output_format == "pandoc"){
    x_t <- knitr::kable(x, format = "pandoc")
  }else{
    x_t <- x %>%
      # x %>%
      # neat() %>%
      knitr::kable(format=output_format) %>%
      kableExtra::kable_styling(
        bootstrap_options = c("striped", "hover", "condensed","responsive"),
        # bootstrap_options = c( "condensed"),
        full_width = F,
        position = "left"
      )
  } 
  return(x_t)
}
# Note: when printing to Word or PDF use `neat(output_format =  "pandoc")`

# functionalize
draw_trends <- function(
  d
  ,selected_measure
){
  # browser()
   target_date <- as.Date("2020-04-15")
  d1 <- d %>%
    filter(measure_type == selected_measure)
  g1 <- d1 %>%
    {
    ggplot(., aes(x=date, y=value, color = measure, group = measure))+
    geom_line(size=4, alpha = .1, color = "black" )+
    geom_line(size=.5)+
    # geom_point(aes(x=target_date), shape = 21, fill = NA,size = 2, color = "black",
               # data = . %>% filter(date == target_date))+
    # geom_vline(xintercept = as.Date("2020-03-11"), linetype="dashed")+
    facet_wrap(vars(!!rlang::sym("measure")), scales = "free_y")+
    scale_y_continuous(labels = scales::comma)+
    scale_x_date(
      breaks = seq(as.Date("2005-01-01"),as.Date("2021-12-31"), by = "12 month"), 
      minor_breaks =  seq(as.Date("2005-01-01"),as.Date("2021-12-31"), by = "3 month"),
      date_labels ="%Y"
    )+  
    labs(
      title = paste0("Income Support Caseload by ", selected_measure)
      ,color = "Metric"
      ,x = "Date"
      ,y = "Number of cases"
      ,caption = "Temporal resolution: 1 month       Minor breaks = 3 months"
    )+
    theme(legend.position = "none")
    }
  g1
}
# Test and Dev
# ds1 %>% draw_trends("Family Composition") 

```


```{r load-data, results='show', message=FALSE, echo =F}
# getwd()
path_file <- "../../data-public/raw/2021-04-16/goasharedcssspqpqara_research-strategiesopen-data00-open-data-asset-packages-2021is-monthlyis-cu.csv"
ds0 <- readr::read_csv(path_file) %>% janitor::clean_names()

```

```{r tweak-data,echo = echo_chunks, results='show', message=message_chunks, echo =F}
ds1 <- ds0 %>%
  mutate(
    date = parse_date_time(ref_date,"m%d%Y") %>% as_date() 
  )
```

# Take away points 

## Observations
- There is a sharp increase in income support caseload around 2009, probably related to 2008 housing bubble, after which the number of clients almost doubled ( from 25,000 to 40,0000).   
- There is a sharp decline in caseload across of [Client Types](#client-type-level),  [Family Composition](family-composition), and [Regions](#regions) after the first quarter of 2020, probably related the introduction of the $2,000 monthly assistance to the unemployed due to COVID in April of 2020.  


## Data issues

- When breaking by [CLient Type Level](#client-type-level), there appears to be a typo: instead of `Client Caseload Total` the value of the `measure` column reads `Client Client Caseload Total` for `10/1/2016`

```{r echo=F, eval=F}
ds1 %>% distinct(measure_type)
ds1 %>% group_by(measure_type) %>% distinct(measure) %>% arrange(measure_type)
```



# Caseload
```{r graph-6, echo = echo_chunks, results='show', message=message_chunks, fig.height = 4}
ds1 %>% draw_trends("Total Caseload") + theme(legend.position = "none")
```

# Client Type Level
```{r graph-3, echo = echo_chunks, results='show', message=message_chunks}
ds1 %>% draw_trends("Client Type Level") + geom_point(data = ds1 %>% filter(measure == "Client Client Caseload Total" ))
cat(" `Client Client Caseload Total` appears to be a typo, so we recode it into `Client Caseload Total` ")
ds1 %>%
  mutate(
    measure = ifelse(measure == "Client Client Caseload Total","Client Caseload Total", measure) 
  ) %>% 
  draw_trends("Client Type Level") 

```



# Family Composition
```{r graph-1, echo = echo_chunks, results='show', message=message_chunks}
ds1 %>% draw_trends("Family Composition")
```



# Regions
```{r graph-7, echo = echo_chunks, results='show', message=message_chunks, fig.height = 8}
ds1 %>% draw_trends("CSS Regions")
```







Session Information {#session-info}
===========================================================================

For the sake of documentation and reproducibility, the current report was rendered in the following environment.  Click the line below to expand.

<details>
  <summary>Environment <span class="glyphicon glyphicon-plus-sign"></span></summary>
```{r session-info, echo=FALSE}
if( requireNamespace("devtools", quietly = TRUE) ) {
  devtools::session_info()
} else {
  sessionInfo()
}
```
</details>

```{r session-duration, echo=FALSE}
report_render_duration_in_seconds <- round(as.numeric(difftime(Sys.time(), report_render_start_time, units="secs")))
```

Report rendered by `r Sys.info()["user"]` at `r strftime(Sys.time(), "%Y-%m-%d, %H:%M %z")` in `r report_render_duration_in_seconds` seconds.
